On the Minimum Core Mass for Giant Planet Formation

### 2013-10-18 tex/draft/draft4.tex
minor corrections secs. 2-4, mostly typos

Full time dependent simulations of giant planet evolution take into account the Kelvin-Helmholtz (KH) contraction of the gaseous envelope. As a result, the atmosphere is no longer in a steady state, but rather it accretes gas as it cools radiatively. In this scenario there are two main sources that contribute to the luminosity evolution of the envelope --- planetesimal accretion and gas contraction. These evolutionary models for core and atmosphere growth include time dependent planetesimal accretion rates that are determined by the distribution of solids in the disk, as well as the planet mass and capture cross-section. This is is contrast with the static models in which the planetesimal accretion rate is usually constant with time. \citet{pollack96} find that the planet growth can be separated into three relatively well defined phases. Phase 1 is dominated by the growth of the core due to runaway planetesimal accretion, and ends when the planet's feeding zone has been depleted of solids. At this stage, planetesimal accretion can no longer balance radiative losses, and the atmosphere cools and contracts. This is phase 2, during which the envelope mass steadily increases, while core growth is no longer significant. Once the atmosphere mass becomes comparable to the core mass, growth is continually accelerated, leading to runaway gas accretion, i.e. phase 3. Phase 2 is found to be much longer than the other two phases (see also \citealt{alibert05}), hence the evolutionary timescale of the atmosphere is set by KH contraction.

@@

Since KH contraction dominates the envelope accretion timescale in many core accretion models, it is useful to study atmospheric growth onto a core of fixed mass (\citealt{ikoma00}, \citealt{pn05}). This regime is the one considered in this work. We explore the atmosphere evolution around fully grown cores of different masses, and determine the minimum core mass to form a giant planet before disk dissipation for a variety of disk conditions and properties of the nebular gas. We note that our models are a special case of the time dependent studies, with the planetesimal accretion rate set to zero. Our neglect of planetesimal accretion allows us to obtain absolute lower limits on the core mass needed to form a gas giant, as additional heat sources limit the ability of the atmosphere to cool and undergo KH contraction.

@@

To model the growth of planetary atmospheres around a solid core, we develop a simplified two layer model for time-dependent atmospheric cooling, i.e. Kelvin-Helmholtz (KH) contraction.  With a convective interior and radiative exterior, this model is motivated by similar models of hot Jupiters \citep{ab06, ym10}.

@@

Our main assumptions are summarized as follows:
\begin{enumerate}
\item The atmosphere is spherically symmetric and remains in hydrostatic balance during its thermal evolution.
\item The core mass and radius are fixed in evolutionary calculations, neglecting ongoing planetesimal or dust accretion.
\item At the planet's Hill radius, the atmospheric temperature and pressure match the conditions of the disk midplane.
\item The only source of planetary luminosity is the gravitational contraction of the atmosphere.  
\item Cooling of the atmosphere is dominated by the convective interior.  Thus the luminosity in the radiative exterior is held spatially constant. %In the radiative zone, luminosity generation is neglected, i.e. the luminosity is held constant.
\item A global cooling model connects independent static solutions into a time-dependent sequence.
\item A polytropic EOS is assumed for simplicity. %AP: define EOS here if not already
\item Dust grains provide the opacity in the radiative zone, which remains cool enough to avoid dust sublimation.
\item Because gas accretion accelerates after the crossover mass, the time to reach the crossover mass, i.e. the crossover time, is a good approximation of the total time to form a gas giant.
\end{enumerate}
The remainder of this section develops our model in more detail.

@@

For a vertically isothermal disk in hydrostatic balance (with no self-gravity), the mid-plane pressure of disk gas is 
\begin{equation}
\label{eq:Pd}
P\di = 6.9 \times 10^{-3} F_\varSigma \sqrt{F_T} \, \aun{10}^{-45/14}~{\rm dyne~cm^2}
%P\di=1.1 \times 10^{-4} F_{\Sigma} \sqrt{F_T} a^{-45/14} \,\, \text{dyne cm}^{-2} %\times \sqrt{m_\ast}
\end{equation}
for a molecular weight of $\mu=2.35$ proton masses and a Solar mass star.  %Changes to the stellar mass (not considered here) would affect both the dynamic mass and, by heating the disk, $T\di$.
%The fiducial pressure is a paltry 7 nanobars.

The (thermodynamically isothermal) sound speed in the disk is
\begin{equation}
c\di = \sqrt{\Rg T\di} = 0.4 \sqrt{F_T} \aun{10}^{3/14} ~\text{km s}^{-1}
\end{equation}  
in terms of the specific gas constant $\Rg$.  The disk scale-height is 
\begin{equation}
H\di = {c\di / \varOmega} = 0.42 \sqrt{F_T}  \, \aun{10}^{9/7} \AU\, .
\end{equation} 
in terms of the Keplerian frequency $\varOmega = \sqrt{G M_\ast/a^3}$ with $G$ the gravitational constant and $M_\ast$ the stellar (in this work Solar) mass.

@@

The characteristic length scales for protoplanetary atmospheres are crucial for choosing boundary conditions and for understanding the validity of  spherical symmetry in a disk of scaleheight $H\di$.  The radius of the solid core:

@@

A planet can bind a dense atmosphere if its escape speed exceeds the sound speed.  This criterion is satisfied inside the Bondi radius
\begin{equation}
\label{eq:RB}
\RB \equiv \frac{G M\pla}{c\di^2} \approx 0.17 \, {\mpn{10}  \, \aun{10}^{3/7} \over F_T} ~\AU
\end{equation}
where the enclosed planet mass, $M\pla = M\co + M_\mathrm{atm}$, includes the core and any atmosphere within the Bondi radius.  The scalings thus use $\mpn{10} \equiv M\pla /(10~M_\oplus)$.

Stellar tides dominate the planet's gravity beyond the Hill radius
\begin{equation}
\label{eq:RHill}
R_{\rm H} = \left(M\pla \over 3 M_\ast \right)^{1/3}a \approx 0.22 \, {\mpn{10}^{1/3} \, \aun{10} }~\AU
\end{equation}
where hydrostatic balance breaks down.  In \Eq{eq:RHill}, $M\pla$ includes mass enclosed within $\RH$.  During the early stages of evolution, $M\pla \sim M\co$, and the core mass can be used to get estimates of both $\RB$ and $\RH$.

The relevant length scales of the atmosphere and disk satisfy the relation $\RB H\di^2 = 3 R_{\rm H}^3$.  The lengthscales are roughly equal at the ``thermal mass'' (e.g., \citealt{menou04})
\begin{equation}
M_{\rm th} > {c\di^{3} \over G \varOmega} \approx 25 \, {F_T^{3/2} \over \sqrt{m_\ast} } \, \aun{10}^{6/7}~ M_\oplus \, .
\end{equation}

In the low mass regime, $M\pla < M_{\rm th}/\sqrt{3}$, the lengthscales order as $\RB< \RH<H\di$ (R06).  In this regime, many studies assume the atmosphere matches the disk conditions at $\RB$.  We however use $\RH$ as the matching radius in both this low mass and other higher mass regimes.  This choice is justified by the fact that, for hydrostatic solutions, the  density at $\RB$ exceeds the disk's background density by an order unity factor (R06).  Nevertheless, since the density change is modest, the choice of outer boundary condition has a similarly modest effect on our results.

@@

Our atmosphere calculations use the standard structure equations of mass conservation, hydrostatic balance, thermal gradients, and energy conservation:
\begin{subeqnarray}
\label{eq:struct}
\frac{dm}{dr}&=&4 \pi r^2 \rho\slabel{eq:structb} \\
\frac{dP}{dr}&=&-\frac{G m}{r^2}\rho \slabel{eq:structa} \\
\frac{dT}{dr}&=&\nabla \frac{T}{P}\frac{dP}{dr}\slabel{eq:structc} \\
\frac{dL}{dr}&=&4 \pi r^2 \rho \left(\epsilon - \left. T {\partial S \over \partial t} \right|_m \right)\slabel{eq:structd}, 
\end{subeqnarray}
\noindent where $r$ is the radial coordinate, $P$, $T$, $\rho$  and $L$ are the gas pressure, temperature, density and luminosity, respectively.  The enclosed mass  at radius $r$ is $m$. \Eq{eq:structc} simply defines the temperature gradient  $\nabla \equiv d \ln T/d \ln P$.  In optically thick radiative zones, radiative diffusion gives a temperature gradient
\begin{equation}
\label{eq:delrad}
\delrad \equiv \frac{3 \kappa P}{64 \pi G m \sigma T^4} L,
\end{equation}
where $\sigma$ is the Stefan-Boltzmann constant. In our models the atmosphere is optically thick throughout the whole radiative region (see \App{opacityan} for an analytic estimate of the optical depth). In convectively unstable regions, efficient convection gives an isentropic temperature gradient with $\nabla = \delad$, the adiabatic gradient

\begin{equation}
\label{eq:delad}
\delad \equiv \Big(\frac{d \ln T}{d \ln P}\Big)_{\rm{ad}}
\end{equation}
According to the Schwarzschild criterion, convective instability occurs when $\delrad > \delad$.  Thus $\nabla = \min(\delrad, \delad)$ sets the temperature gradient.

@@

To solve the equation set (\ref{eq:struct}) an equation of state (EOS) is required for closure. In our study, we adopt an ideal gas law with a polytropic EOS 
\begin{subeqnarray}
P &=& \rho \Rg T \, ,\slabel{eq:idealgas} \\
P &=&K \rho^{\gamma} \, , \slabel{eq:polyEOS}%\equiv K \rho^{1/(1-\delad)}, 
\end{subeqnarray}
where $K$ is the adiabatic constant. The adiabatic index  $\gamma = 1/(1 - \delad)$.   An ideal monatomic gas has $\delad = 2/5$.  This work uses $\delad=2/7$ for an ideal diatomic gas.  While our reference mean molecular weight ($\mu = 2.35$ proton masses) includes Helium, we ignore Helium's effect on the EOS, which is already greatly simplified. The second law or thermodynamics gives the relative entropy as
\begin{equation}
S = \Rg \ln \left(T^{1/\delad} \over P \right) 
\end{equation} 
eliminating the need for $K$.

Boundary conditions must be satisfied at both the base and the top of the atmosphere with $m(R\co) = M\co$, $T(\RH) = T\di$ and $P(\RH) = P\di$.  In principle our solutions describe atmospheres with $L(R\co) = 0$. In practice, since we do not directly integrate \Eq{eq:structd} we need not directly impose this boundary condition, as described in \S\ref{sec:twolayer}.

@@

This section describes the global energy balance of a planet embedded in a gas disk, or more generally any spherical, hydrostatic object in pressure equilibrium with a background medium.  The total atmospheric energy includes gravitational and internal energies, $E = E_G + U$:
\begin{subeqnarray}
E_G&=&-\int_{M\co}^M \frac{G m}{r} dm \, , \label{eq:Eg} \\
U&=&\int_{M\co}^M u dm \, .\slabel{eq:U}
\end{subeqnarray}
The specific internal energy $u = C_V T = \Rg (\delad^{-1} -1) T$ for a polytropic EOS.  For a star or coreless planet, $M\co = 0$.

@@

A protoplanetary atmosphere embedded in a gas disk lacks a free surface.  For objects without a free surface (or interior to a free surface), the full energy equation, 
\begin{equation}
\label{eq:coolingglobal}
L_M=L\co+\Gamma-\dot{E}+e_M \dot{M} - P_M \left. \frac{\partial V_M}{\partial t}\right|_M \, ,
\end{equation}
acquires surface terms as derived in  \App{sec:globalderiv}.  The surface can be at any mass level $M$, where the instantaneous radius is $R$.  Other surface quantities are labelled by $M$ subscripts.  The energy accreted across the surface is given by the specific energy, $e_M = u_M-G M/R$, and the mass accretion rate of gas, $\dot{M}$.  The work done by the surface is $P_M \partial V_M/ \partial t$, with the partial derivative performed at fixed mass.  %This generalized energy equation applies on any spherical shell where hydrostatic balance holds.  The choice of the mass shell $M$\footnote{\Eq{eq:coolingglobal} also applies in the interior of objects with a free surface.  The work term, $P_M \p V_M/\p t$, vanishes at the free surface.  The accretion energy, $e_M \dot{M}$,  vanishes for a truly isolated object, but would be included to account for accretion onto (or mass loss from) an otherwise free surface.}   Thus $M$ no longer refers to a uniquely defined total mass, but to the chosen mass level, where the instantaneous radius is $R$.  Values on this shell are labelled by $M$ subscripts.

For static solutions, which are not the focus of this Paper, the surface terms (and also $\dot{E})$ vanish.  Static solutions are valid when imposed heat sources, i.e.\ $L\co$ and $\Gamma$, exceed the atmospheric losses.  Quantitatively, static solutions apply when the KH timescale,
\begin{equation}
\tau_{\rm KH} \sim {|E| \over L_M}, 
\end{equation} 
is shorter than the actual evolutionary timescale.  Thus, $\tau_{\rm KH}$, which our models calculate, gives strong lower limits on the time to form giant planets by core accretion.

@@

To simplify our calculations of atmospheric contraction we use a two layer model with a bottom convective region and an upper radiative layer.   The existence of such a structure is well known from previous studies (e.g., R06) and can be readily understood.  Before the protoplanetary atmosphere can cool, it has the entropy of the disk.  As the atmosphere cools the deep interior remains convective.  Convective interiors are a common feature of low mass cool objects (brown dwarfs and planets) that results from the behavior of $\delrad$ for realistic opacity laws.  However the entropy of the deep interior decreases as the atmosphere cools.  A region of outwardly increasing entropy, i.e.\ a radiative layer, is required to connect the convective interior to the disk.  A more complicated structure, with radiative windows in the convection zone, is possible as discussed in \S\ref{sec:opEOS}.

In convective regions, the adiabatic structure is independent of luminosity and can be calculated without local energy balance, \Eq{eq:structd}.  Thus for fully convective objects, a cooling sequence can be established by connecting a series of adiabatic solutions using a global energy equation, $L_M = -\dot{E}$ or \Eq{eq:coolingstar}.  Such methods are commonly used for their computational efficiency and are sometime referred to as ``following the adiabats," since the steady state solutions evolve in order of decreasing entropy \citep{marleau13}.

In the radiative zone,  local energy balance, \Eq{eq:structd}, does affect the atmospheric structure.  We proceed by assuming that the majority of energy is lost from the convective interior, and thus the luminosity can be treated as constant in the outer radiative zone.  With this approximation we can construct solutions from equations (\ref{eq:struct}a -- c) that ``follow the mass," i.e.\ gradually increase the atmospheric mass.  We then use the global energy balance, \Eq{eq:coolingglobal} to place these solutions in a cooling sequence.  The validity of neglecting luminosity generation in the radiative zone can be checked \emph{a posteriori}.

To make a single atmosphere model (indexed by $i$) we choose a planet mass $M_i$.  At the outer boundary, at $\RH(M_i)$, the temperature and pressure are set to the disk values.  To integrate equations (\ref{eq:struct}a--c), the luminosity is required to compute $\delrad$.  The correct value of the luminosity is not known in advance, and is the eigenvalue of the problem.  The boundary conditions can only be satisfied for the correct value of the luminosity eigenvalue, which we find by the shooting method.  Specifically, we alter the luminosity until the integrated value of mass at the core, $m(R\co)$, matches the actual core mass, $M\co$.

@@

This section develops the analytic version of our two-layer model, which is useful for predicting and explaining the results of numerical evolution models.  To obtain our analytic results, we neglect self-gravity.   As shown in Section \ref{sec:KH}, self-gravity becomes quantitatively important at surprisingly low atmospheric masses, $M_{\rm atm} \gtrsim 0.2 M\co$.  Thus the analytic model cannot accurately model the later stages of evolution.  Nevertheless, this model does explain general trends, including the dependence on opacity and disk temperature and pressure.

The analytic model also assumes that the upper radiative radiative layer is thick, with a pressure depth $P\cb \gg P\di$.  This approximation limits the ability to study the initial cooling of very low mass atmospheres with thin outer radiative zones.  However since early evolution is quite rapid, this approximation negligibly affects total cooling times and critical core masses, our main foci.

\subsection{Two Layer Structure}
To apply the two layer cooling model analytically, we require expressions for the atmospheric structure.  Conditions at the RCB are crucial as they set the interior adiabat and the radiative losses. (Recall that luminosity generation in the radiative zone is neglected.)  We express the temperature and pressure of the RCB, at the radius $R\cb$, as 
\begin{subeqnarray}\label{eq:cb2}
T\cb &=& \chi T\di \slabel{eq:Tcb} \\
P\cb &=& \theta P_{\rm d} e^{R_{\rm B}/R\cb} \, .\slabel{eq:PcbRcb}
\end{subeqnarray}

The leading constants would be unity, $\chi = \theta = 1$, if the radiative zone were replaced by an isothermal layer.  In practice deviations from unity are modest.    Standard radiative structure calculations (see \App{RCBcorr} for details) give 
\begin{equation}
\label{eq:chi}
\chi \simeq \Big(1-\frac{\delad}{\nabla_{\infty}} \Big)^{-\frac{1}{4-\beta}} \simeq 1.53 \, ,
\end{equation}
for $P\cb \gg P\di$, our regime of interest, and with the radiative temperature gradient at infinite depth, $\nabla_\infty = 1/2$, for our dust opacity.  The radiative zones are thus nearly isothermal, as also found by R06.  While  a simple closed form expression is not possible, the factor $\theta \simeq 0.556$ for our parameters.   
%According to \Eq{eq:PcbRcb} our pressure criterion also gives $R\cb \ll \RB$, but the inequality is logarithmically weaker.

@@

Both $t_{\rm co}$ and $\MC$ are too large to be interesting, or to be correct based on previous results and the numerical results in this Paper.  The main reason for this discrepancy is the neglect of self-gravity, as shown in Section \ref{sec:KH}.  The magnitude of the discrepancy may be surprising as one might expect self-gravity to be a modest factor for $M_{\rm atm} \leq M\co$.  Clearly this seemingly reasonable expectation can be misleading, an interesting result in itself.

>>>>

Full time-dependent simulations of giant planet evolution take into account the Kelvin-Helmholtz (KH) contraction of the gaseous envelope. As a result, the atmosphere is no longer in a steady state, but rather it accretes gas as it cools radiatively. In this scenario there are two main sources that contribute to the luminosity evolution of the envelope --- planetesimal accretion and gas contraction. These evolutionary models for core and atmosphere growth include time-dependent planetesimal accretion rates that are determined by the distribution of solids in the disk, as well as the planet mass and capture cross-section. This is is contrast with the static models in which the planetesimal accretion rate is usually constant with time. \citet{pollack96} find that the planet growth can be separated into three relatively well defined phases. Phase 1 is dominated by the growth of the core due to runaway planetesimal accretion, and ends when the planet's feeding zone has been depleted of solids. At this stage, planetesimal accretion can no longer balance radiative losses, and the atmosphere cools and contracts. This is phase 2, during which the envelope mass steadily increases, while core growth is no longer significant. Once the atmosphere mass becomes comparable to the core mass, growth is continually accelerated, leading to runaway gas accretion, i.e. phase 3. Phase 2 is found to be much longer than the other two phases (see also \citealt{alibert05}), hence the evolutionary timescale of the atmosphere is set by KH contraction.

@@

Since KH contraction dominates the envelope accretion timescale in many core accretion models, it is useful to study atmospheric growth onto a core of fixed mass (\citealt{ikoma00}, \citealt{pn05}). This regime is the one considered in this work. We explore the atmosphere evolution around fully grown cores of different masses, and determine the minimum core mass to form a giant planet before disk dissipation for a variety of disk conditions and properties of the nebular gas. We note that our models are a special case of the time-dependent studies, with the planetesimal accretion rate set to zero. Our neglect of planetesimal accretion allows us to obtain absolute lower limits on the core mass needed to form a gas giant, as additional heat sources limit the ability of the atmosphere to cool and undergo KH contraction.

@@

To model the growth of planetary atmospheres around a solid core, we develop a simplified two layer model for time-dependent atmospheric cooling, i.e. KH contraction.  With a convective interior and radiative exterior, this model is motivated by similar models of hot Jupiters \citep{ab06, ym10}.

@@

Our main assumptions are summarized as follows:
\begin{enumerate}
\item The atmosphere is spherically symmetric and remains in hydrostatic balance during its thermal evolution.
\item The core mass and radius are fixed in evolutionary calculations, neglecting ongoing planetesimal or dust accretion.
\item At the planet's Hill radius, the atmospheric temperature and pressure match the conditions of the disk midplane.
\item The only source of planetary luminosity is the gravitational contraction of the atmosphere.  
\item Cooling of the atmosphere is dominated by the convective interior.  Thus the luminosity in the radiative exterior is held spatially constant. %In the radiative zone, luminosity generation is neglected, i.e. the luminosity is held constant.
\item A global cooling model connects independent static solutions into a time-dependent sequence.
\item A polytropic equation of state (EOS) is assumed for simplicity. %AP: define EOS here if not already
\item Dust grains provide the opacity in the radiative zone, which remains cool enough to avoid dust sublimation.
\item Because gas accretion accelerates after the crossover mass, the time to reach the crossover mass, i.e. the crossover time, is a good approximation of the total time to form a gas giant.
\end{enumerate}
The remainder of this section develops our model in more detail.

@@

For a vertically isothermal disk in hydrostatic balance (with no self-gravity), the mid-plane pressure of disk gas is 
\begin{equation}
\label{eq:Pd}
P\di = 6.9 \times 10^{-3} F_\varSigma \sqrt{F_T} \, \aun{10}^{-45/14}~{\rm dyn~cm^{-2}}
%P\di=1.1 \times 10^{-4} F_{\Sigma} \sqrt{F_T} a^{-45/14} \,\, \text{dyne cm}^{-2} %\times \sqrt{m_\ast}
\end{equation}
for a molecular weight of $\mu=2.35$ proton masses and a Solar mass star.  %Changes to the stellar mass (not considered here) would affect both the dynamic mass and, by heating the disk, $T\di$.
%The fiducial pressure is a paltry 7 nanobars.

The (thermodynamically isothermal) sound speed in the disk is
\begin{equation}
c\di = \sqrt{\Rg T\di} = 0.4 \sqrt{F_T} \aun{10}^{3/14} ~\text{km s}^{-1}
\end{equation}  
in terms of the specific gas constant $\Rg$.  The disk scale height is 
\begin{equation}
H\di = {c\di / \varOmega} = 0.42 \sqrt{F_T}  \, \aun{10}^{9/7} \AU\, .
\end{equation} 
in terms of the Keplerian frequency $\varOmega = \sqrt{G M_\ast/a^3}$, with $G$ the gravitational constant and $M_\ast$ the stellar (in this work Solar) mass.

@@

The characteristic length scales for protoplanetary atmospheres are crucial for choosing boundary conditions and for understanding the validity of  spherical symmetry in a disk of scale height $H\di$.  The radius of the solid core is

@@

A planet can bind a dense atmosphere if its escape velocity exceeds the sound speed.  This criterion is satisfied inside the Bondi radius
\begin{equation}
\label{eq:RB}
\RB \equiv \frac{G M\pla}{c\di^2} \approx 0.17 \, {\mpn{10}  \, \aun{10}^{3/7} \over F_T} ~\AU
\end{equation}
where the enclosed planet mass, $M\pla = M\co + M_\mathrm{atm}$, includes the core and any atmosphere within the Bondi radius.  The scalings thus use $\mpn{10} \equiv M\pla /(10~M_\oplus)$.

Stellar tides dominate the planet's gravity beyond the Hill radius
\begin{equation}
\label{eq:RHill}
R_{\rm H} = \left(M\pla \over 3 M_\ast \right)^{1/3}a \approx 0.22 \, {\mpn{10}^{1/3} \, \aun{10} }~\AU ,
\end{equation}
where hydrostatic balance breaks down.  In \Eq{eq:RHill}, $M\pla$ includes mass enclosed within $\RH$.  During the early stages of evolution, $M\pla \sim M\co$, and the core mass can be used to get estimates of both $\RB$ and $\RH$.

The relevant length scales of the atmosphere and disk satisfy the relation $\RB H\di^2 = 3 R_{\rm H}^3$.  The length scales are roughly equal at the ``thermal mass'' (e.g., \citealt{menou04})
\begin{equation}
M_{\rm th} > {c\di^{3} \over G \varOmega} \approx 25 \, {F_T^{3/2} \over \sqrt{m_\ast} } \, \aun{10}^{6/7}~ M_\oplus \, .
\end{equation}

In the low mass regime, $M\pla < M_{\rm th}/\sqrt{3}$, the length scales order as $\RB< \RH<H\di$ (\citealt{rafikov06}, hereafter R06).  In this regime, many studies assume the atmosphere matches the disk conditions at $\RB$.  We, however, use $\RH$ as the matching radius in both this low mass and other higher mass regimes.  This choice is justified by the fact that, for hydrostatic solutions, the  density at $\RB$ exceeds the disk's background density by an order unity factor (R06).  Nevertheless, since the density change is modest, the choice of outer boundary condition has a similarly modest effect on our results.

@@

Our atmosphere calculations use the standard structure equations of mass conservation, hydrostatic balance, thermal gradients, and energy conservation:
\begin{subeqnarray}
\label{eq:struct}
\frac{dm}{dr}&=&4 \pi r^2 \rho\slabel{eq:structb} \\
\frac{dP}{dr}&=&-\frac{G m}{r^2}\rho \slabel{eq:structa} \\
\frac{dT}{dr}&=&\nabla \frac{T}{P}\frac{dP}{dr}\slabel{eq:structc} \\
\frac{dL}{dr}&=&4 \pi r^2 \rho \left(\epsilon - \left. T {\partial S \over \partial t} \right|_m \right)\slabel{eq:structd}, 
\end{subeqnarray}
\noindent where $r$ is the radial coordinate, and $P$, $T$, $\rho$  and $L$ are the gas pressure, temperature, density and luminosity, respectively.  The enclosed mass  at radius $r$ is $m$. \Eq{eq:structc} simply defines the temperature gradient  $\nabla \equiv d \ln T/d \ln P$.  In optically thick radiative zones, radiative diffusion gives a temperature gradient
\begin{equation}
\label{eq:delrad}
\delrad \equiv \frac{3 \kappa P}{64 \pi G m \sigma T^4} L,
\end{equation}
where $\sigma$ is the Stefan-Boltzmann constant. In our models the atmosphere is optically thick throughout the whole radiative region (see \App{opacityan} for an analytic estimate of the optical depth). In convectively unstable regions, efficient convection gives an isentropic temperature gradient with $\nabla = \delad$, the adiabatic gradient

\begin{equation}
\label{eq:delad}
\delad \equiv \Big(\frac{d \ln T}{d \ln P}\Big)_{\rm{ad}}.
\end{equation}
According to the Schwarzschild criterion, convective instability occurs when $\delrad > \delad$.  Thus $\nabla = \min(\delrad, \delad)$ sets the temperature gradient.

@@

To solve the equation set (\ref{eq:struct}), an EOS is required for closure. In our study, we adopt an ideal gas law with a polytropic EOS 
\begin{subeqnarray}
P &=& \rho \Rg T \, ,\slabel{eq:idealgas} \\
P &=&K \rho^{\gamma} \, , \slabel{eq:polyEOS}%\equiv K \rho^{1/(1-\delad)}, 
\end{subeqnarray}
where $K$ is the adiabatic constant. The adiabatic index is  $\gamma = 1/(1 - \delad)$.  This work uses $\delad=2/7$ for an ideal diatomic gas (by comparison, an ideal monatomic gas has $\delad = 2/5$).  While our reference mean molecular weight ($\mu = 2.35$ proton masses) includes Helium, we ignore Helium's effect on the EOS, which is already greatly simplified. The second law or thermodynamics gives the relative entropy as
\begin{equation}
S = \Rg \ln \left(T^{1/\delad} \over P \right) ,
\end{equation} 
eliminating the need for $K$.

Boundary conditions must be satisfied at both the base and the top of the atmosphere with $m(R\co) = M\co$, $T(\RH) = T\di$ and $P(\RH) = P\di$.  In principle, our solutions describe atmospheres with $L(R\co) = 0$. In practice, since we do not directly integrate \Eq{eq:structd} we need not directly impose this boundary condition, as described in \S\ref{sec:twolayer}.

@@

This section describes the global energy balance of a planet embedded in a gas disk, or more generally any spherical, hydrostatic object in pressure equilibrium with a background medium.  The total atmospheric energy includes gravitational and internal energies, $E = E_G + U$, with
\begin{subeqnarray}
E_G&=&-\int_{M\co}^M \frac{G m}{r} dm \, , \label{eq:Eg} \\
U&=&\int_{M\co}^M u dm \, .\slabel{eq:U}
\end{subeqnarray}
The specific internal energy is $u = C_V T = \Rg (\delad^{-1} -1) T$ for a polytropic EOS.  For a star or coreless planet, $M\co = 0$.

@@

A protoplanetary atmosphere embedded in a gas disk lacks a free surface.  For objects without a free surface (or interior to a free surface), the full energy equation, 
\begin{equation}
\label{eq:coolingglobal}
L_M=L\co+\Gamma-\dot{E}+e_M \dot{M} - P_M \left. \frac{\partial V_M}{\partial t}\right|_M \, ,
\end{equation}
acquires surface terms as derived in  \App{sec:globalderiv}.  The surface can be at any mass level $M$, where the instantaneous radius is $R$.  Other surface quantities are labeled by $M$ subscripts.  The energy accreted across the surface is given by the specific energy, $e_M = u_M-G M/R$, and the mass accretion rate of gas, $\dot{M}$.  The work done by the surface is $P_M \partial V_M/ \partial t$, with the partial derivative performed at fixed mass.  %This generalized energy equation applies on any spherical shell where hydrostatic balance holds.  The choice of the mass shell $M$\footnote{\Eq{eq:coolingglobal} also applies in the interior of objects with a free surface.  The work term, $P_M \p V_M/\p t$, vanishes at the free surface.  The accretion energy, $e_M \dot{M}$,  vanishes for a truly isolated object, but would be included to account for accretion onto (or mass loss from) an otherwise free surface.}   Thus $M$ no longer refers to a uniquely defined total mass, but to the chosen mass level, where the instantaneous radius is $R$.  Values on this shell are labelled by $M$ subscripts.

For static solutions, which are not the focus of this paper, the surface terms (and also $\dot{E})$ vanish.  Static solutions are valid when imposed heat sources, i.e.\ $L\co$ and $\Gamma$, exceed the atmospheric losses.  Quantitatively, static solutions apply when the KH timescale,
\begin{equation}
\tau_{\rm KH} \sim {|E| \over L_M}, 
\end{equation} 
is shorter than the actual evolutionary timescale.  Thus, $\tau_{\rm KH}$, which our models calculate, gives strong lower limits on the time to form giant planets by core accretion.

@@

To simplify our calculations of atmospheric contraction we use a two layer model with a bottom convective region and an upper radiative layer.   The existence of such a structure is well known from previous studies (e.g., R06) and can be readily understood.  Before the protoplanetary atmosphere can cool, it has the entropy of the disk.  As the atmosphere cools, the deep interior remains convective.  Convective interiors are a common feature of low mass cool objects (brown dwarfs and planets) that results from the behavior of $\delrad$ for realistic opacity laws.  However, the entropy of the deep interior decreases as the atmosphere cools.  A region of outwardly increasing entropy, i.e.\ a radiative layer, is required to connect the convective interior to the disk.  A more complicated structure, with radiative windows in the convection zone, is possible as discussed in \S\ref{sec:op}.

In convective regions, the adiabatic structure is independent of luminosity and can be calculated without local energy balance, \Eq{eq:structd}.  Thus, for fully convective objects, a cooling sequence can be established by connecting a series of adiabatic solutions using a global energy equation, $L_M = -\dot{E}$ or \Eq{eq:coolingstar}.  Such methods are commonly used for their computational efficiency and are sometime referred to as ``following the adiabats," since the steady state solutions evolve in order of decreasing entropy \citep{marleau13}.

In the radiative zone,  local energy balance, \Eq{eq:structd}, does affect the atmospheric structure.  We proceed by assuming that the majority of energy is lost from the convective interior, and thus the luminosity can be treated as constant in the outer radiative zone.  With this approximation we can construct solutions from Equations (\ref{eq:struct}a -- c) that ``follow the mass," i.e.\ gradually increase the atmospheric mass.  We then use global energy balance, \Eq{eq:coolingglobal}, to place these solutions in a cooling sequence.  The validity of neglecting luminosity generation in the radiative zone can be checked \emph{a posteriori}.

To make a single static atmosphere model (indexed by $i$), we choose a planet mass $M_i$.  At the outer boundary, at $\RH(M_i)$, the temperature and pressure are set to the disk values.  To integrate Equations (\ref{eq:struct}a--c), the luminosity is required to compute $\delrad$.  The correct value of the luminosity is not known in advance, and is the eigenvalue of the problem.  The boundary conditions can only be satisfied for the correct value of the luminosity eigenvalue, which we find by the shooting method.  Specifically, we alter the luminosity until the integrated value of mass at the core, $m(R\co)$, matches the actual core mass, $M\co$.

@@

This section develops the analytic version of our two-layer model, which is useful for predicting and explaining the results of numerical evolution models.  To obtain our analytic results, we neglect self-gravity.   As shown in  \S\ref{sec:KH}, self-gravity becomes quantitatively important at surprisingly low atmospheric masses, $M_{\rm atm} \gtrsim 0.2 M\co$.  Thus the analytic model cannot accurately model the later stages of evolution.  Nevertheless, this model does explain general trends, including the dependence on opacity and disk temperature and pressure.

The analytic model also assumes that the upper radiative radiative layer is thick, with a pressure depth $P\cb \gg P\di$.  This approximation limits the ability to study the initial cooling of very low mass atmospheres with thin outer radiative zones.  However, since early evolution is quite rapid, this approximation negligibly affects total cooling times and critical core masses, our main foci.

\subsection{Two Layer Structure}
To apply the two layer cooling model analytically, we require expressions for the atmospheric structure.  Conditions at the RCB are crucial as they set the interior adiabat and the radiative losses. (Recall that luminosity generation in the radiative zone is neglected.)  We express the temperature and pressure of the RCB, at the radius $R\cb$, as 
\begin{subeqnarray}\label{eq:cb2}
T\cb &=& \chi T\di \slabel{eq:Tcb} \\
P\cb &=& \theta P_{\rm d} e^{R_{\rm B}/R\cb} \, .\slabel{eq:PcbRcb}
\end{subeqnarray}
The leading constants would be unity, $\chi = \theta = 1$, if the radiative zone were replaced by an isothermal layer.  In practice, deviations from unity are modest.    Standard radiative structure calculations (see \App{RCBcorr} for details) give 
\begin{equation}
\label{eq:chi}
\chi \simeq \Big(1-\frac{\delad}{\nabla_{\infty}} \Big)^{-\frac{1}{4-\beta}} \simeq 1.53 \, ,
\end{equation}
for $P\cb \gg P\di$, our regime of interest, and with the radiative temperature gradient at infinite depth, $\nabla_\infty = 1/2$, for our dust opacity.  The radiative zones are thus nearly isothermal, as also found by R06.  While  a simple closed form expression is not possible, the factor $\theta \simeq 0.556$ for our parameters.   
%According to \Eq{eq:PcbRcb} our pressure criterion also gives $R\cb \ll \RB$, but the inequality is logarithmically weaker.

@@

Both $t_{\rm co}$ and $\MC$ are too large to be interesting, or to be correct based on previous results and the numerical results in this paper.  The main reason for this discrepancy is the neglect of self-gravity, as shown in Section \ref{sec:KH}.  The magnitude of the discrepancy may be surprising as one might expect self-gravity to be a modest factor for $M_{\rm atm} \leq M\co$.  Clearly this seemingly reasonable expectation can be misleading, an interesting result in itself.
